---
title: Number lines in code blocks with Rmarkdown
author: Atusy
date: |
  Created on 2019-04-18. \
  Updated on `r Sys.Date()`.
output: 
  # works
  html_document: 
    highlight: pygment
    toc: true
    toc_float: true
    number_sections: true
  pdf_document: default
  # not works
  html_notebook: 
    highlight: pygment
  word_document: default
---

**Before reading**

- For quick start, go "[Autonumbering]".
- A source and output examples in variety of formats are on GitHub [`atusy/rmd-line-num`](https://github.com/atusy/rmd-line-num).
- Ask me questions on [GitHub Issues](https://github.com/atusy/rmd-line-num/issues) or on Twitter ([\@Atsushi776](https://twitter.com/Atsushi776))

# Number Pandoc's fenced code

Pandoc has an official support to number lines on fenced code by giving a class attribute `numberLines` (<https://www.pandoc.org/MANUAL.html#extension-fenced_code_attributes>).
The document also introduces that
`lineAnchors` class makes line numbers be clickable, and
`startFrom` arugment supports specify starting number.

Note that class attributes require `.` before the class name.

````{.numberLines}
```{.r .numberLines .lineAnchors startFrom="11"}
x <- rnorm(10)
mean(x)
```
````

becomes

```{.r .numberLines .lineAnchors startFrom="11"}
x <- rnorm(10)
mean(x)
```

This is enough for `rmarkdown::pdf_document`.

For `html_document`, we also need to specify `highlight` other than `default` in YAML front matter[^highlight]^,^ [^niszet]. See `?rmarkdown::html_document` for available `highlight` options.

[^highlight]: "Chunk numberLines hook]" on Rpubs figured out requirement of `pygment` (<https://rpubs.com/Thell/numberLines>).

[^niszet]: [\@niszet0]() mentioned that not only `pygment` but also `highlight` other than `default` are acceptable (<https://niszet.hatenablog.com/entry/2019/04/18/194246> in Japanese).

```{.yaml .numberLines}
---
title: Line numbers with Rmarkdown documents
output: 
  html_document:
    highlight: pygment
  pdf_document: default
---
```



# Number Rmarkdown's chunks and outputs

The above success infers a success in chunks of Rmarkdown.

## Numbering chunks

For code chunks of Rmarkdown documents,
`numberLines` class can be given by assigning `class.source = "numberLines"` as a chunk option[^yihui]. 
You may also add `lineAnchors` demilited by space (`"numberLines lineAnchors`") or as another element of vector (`c("numberLines", "lineAnchors")`) [^niszet-tweet].
Unfortunately, I find no way to add `startFrom="integer"`.

[^yihui]: See "Tex Results" section of "Chunk options and package options" by Yi Hui (<https://yihui.name/knitr/options/#text-results>)

[^niszet-tweet]: [\@niszet](https://twitter.com/niszet0) mentioned me on Twitter <https://twitter.com/niszet0/status/1118837863200591872>

Again, you need to specify `highlight` for `html_document`.

Thus,

````{.numberLines}
---
title: Line numbers with Rmarkdown documents
output: 
  html_document:
    highlight: pygment
---

```{`r "r"`, class.source = "numberLines lineAnchors"}
x <- seq(10)
mean(x)
```
````

becomes

```{r, class.source = "numberLines lineAnchors"}
x <- seq(10)
mean(x)
```

Great again!!


Unfortunately, line numbering does not work on `rmarkdown::word_document`.

It does not work properly on `rmarkdown::html_notebook` and `blogdown::html_page` as well.
I guess some tricks required in CSS or JS.

## Numbering outputs

You may also want to number lines on outputs by `class.output = "numberLines"` with or without `lineAnchors` just like input.
However, this changes background colors to gray.

```{r, class.source = "numberLines", class.output = "numberLines", echo = FALSE}
x <- seq(10)
mean(x)
```

If output format is html, css will help.

Before adding `class.output = "numberLines"`, an output in html is

```{.html .numberLines}
<pre><code>## [1] 5.5</code></pre>
```

After adding it, the output in html becomes

```{.html .numberLines}
<div class="sourceCode" id="cb7"><pre class="sourceCode numberSource numberLines"><code class="sourceCode"><a class="sourceLine" id="cb7-1" title="1">## [1] 5.5</a></code></pre></div>
```

When `chunkout` class is further added, the output in html becomes

```{.html .numberLines}
<div class="sourceCode" id="cb10"><pre class="sourceCode numberSource numberLines chunkout"><code class="sourceCode"><a class="sourceLine" id="cb10-1" title="1">## [1] 5.5</a></code></pre></div>
```

You can see `chunkout` class is added to `pre` tag.

So, lets modify css with

```{css, class.source = "numberLines"}
div.sourceCode pre.chunkout {
  background: white;
}
```

and you'll be happy, right?

```{r, class.source = "numberLines", class.output = c("numberLines", "chunkout"), echo = FALSE}
x <- seq(10)
mean(x)
```

Sorry I do not support pdf because $LaTeX$ kills me.

# Autonumbering

- Super easy by a following template and edit after `<!-- Start your body -->`.
- Disable numbering by `class.source = NULL, class.output = NULL`.
- Note autonumbering does not work on Pandoc's fenced code [^pandoc].
  
  [^pandoc]: The document indicates a following YAML front matter should work, but doesn't (<https://www.pandoc.org/MANUAL.html#reader-options>).

  ```{.yaml .numberLines}
  ---
  output:
    html_document:
      highlight: pygment
      pandoc_args:
        - --indented-code-classes
        - lineNumbers
  ---
  ```

## Input with a template

````{.numberLines}
---
output:
  html_document:
    highlight: pygment
  pdf_document: default
---

```{`r "r"` setup, include = FALSE}
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors", 
  class.output = c("numberLines lineAnchors chunkout") 
)

# Add some arbitrary setup codes
```


```{`r "css"`, echo = FALSE}
div.sourceCode pre.chunkout {
  background: white;
}
```

<!-- Start your body -->

**Numbered**

```{`r "r"`}
x <- seq(10)
mean(x)
```

**Unnumbered**

```{`r "r"`, class.source = NULL, class.output = NULL}
x <- seq(10)
mean(x)
```

````

## Output from the template

**Numbered**

```{r, include = FALSE}
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors", 
  class.output = "numberLines lineAnchors chunkout"
)
```

```{r}
x <- seq(10)
mean(x)
```

**Unnumbered**

```{r, class.source = NULL, class.output = NULL}
x <- seq(10)
mean(x)
```

# Enjoy !