---
title: Marginal distributions with ggplot2 and patchwork
author: ~
date: '2018-11-02'
slug: marginal-patchwork
categories: [Tech]
tags: [R, R-ggplot2, R-patchwork]
---

```{r setup, include = FALSE}
library(ggplot2)
library(patchwork)
```

patchwork packages is a great tool to assemble ggplot2 object.

I made a Japanese [slide](https://atusy.github.io/presentation/tokyor073/tokyor073-multi-ggplot2.html?raw#/patchwork-) to introduce patchwork package in Tokyo R meeting.

Here, I tried to produce marginal plots, but failed because I was using plot arithmetics (`|` and `/`).

```{r}
library(ggplot2)
library(patchwork)
xy <- ggplot(mtcars, aes(wt, mpg)) + geom_point()
x <- ggplot(mtcars, aes(wt)) + geom_histogram(bins = 30)
y <- ggplot(mtcars, aes(mpg)) + geom_histogram(bins = 30) + coord_flip()
(x | plot_spacer()) / (xy | y)
```

I just found that `wrap_plots()` helps.

```{r}
wrap_plots(x, plot_spacer(), xy, y, nrow = 2)
```

However, plots need to share `xlim` and `ylim` using `coord_cartesian()` and `coord_flip()`.

`xlim()`, `ylim()`, are not good choice because they may change binwidths of histograms.

```{r, message = FALSE}
wrap_plots(
  x + coord_cartesian(xlim = c(1, 6)), 
  plot_spacer(), 
  xy + coord_cartesian(xlim = c(1, 6), ylim = c(10, 35)), 
  y + coord_flip(xlim = c(10, 35)), 
  nrow = 2
)
```

Adjusting `theme()` and `wrap_plots(widths =, heights =)` will make much more beautiful marginal plots.

```{r, message = FALSE}
theme_marginal_x <- theme(axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank())
theme_marginal_y <- theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())
wrap_plots(
  x + coord_cartesian(xlim = c(1, 6)) + theme_marginal_x, 
  plot_spacer(), 
  xy + coord_cartesian(xlim = c(1, 6), ylim = c(10, 35)), 
  y + coord_flip(xlim = c(10, 35)) + theme_marginal_y, 
  nrow = 2,
  widths = c(1, 0.5),
  heights = c(0.5, 1)
)
```


