---
title: Benchmark chunks in Rmd
author: ~
date: '2018-08-18'
slug: time-each-chunk
categories: [Tech]
tags: [R, Rmarkdown, R-knitr]
output:
  blogdown::html_page:
    toc: yes
---

```{r setup, include = FALSE}
library(pacman)
p_load_gh("rkrug/plantuml")
p_load(knitr)
```


In order to benchmark chunks, 
I implemented a 
[hook](https://yihui.name/knitr/hooks/)
for Rmd's chunk which works as if Jupyter Notebook's 
[`%%timeit`](https://jakevdp.github.io/PythonDataScienceHandbook/01.07-timing-and-profiling.html)
command.

# Implementation

A simple benchmark is done by saving `proc.time()` before running code and subtract the value from new `proc.time()` after code runs.

```{r}
t <- proc.time()['elapsed']
invisible(runif(100))
proc.time()['elapsed'] - t
```

However, hooks are functions, and 
bodies of functions are evaluated in different environments every time.

```{r}
f <- function() environment()
c(f(), f())
```

Thus, an implementing hook function has to know environment to save `proc.time()`, and to call the saved `proc.time()`.

Fortunately, the hook with `envir` argument enables it because `envir` captures an environment in which a code chunk is evaluated.

In addition, the hook needs to know if they are called before or after a code chunk, which is possible by using `before` argument.

Finally, the hook is implemented as below.

```{r}
library(knitr)
knit_hooks$set(
  time = function(before, envir) {
    t <- proc.time()['elapsed']
    if(before) {
      envir$.elapsed <- t
    } else {
      paste(signif(t - envir$.elapsed), 'sec.')
    }
  }
)
```

Use this hook by setting `time = TRUE` as a chunk option.

FYI, returning value must be a character to be printed, and 
`signif()` is used to focus on significant digits.

# Test

```{r, include = FALSE}
x1 <- "```{r, time = TRUE}"
x2 <- "head(x)"
x3 <- "```"
```

    `r x1`
    `r x2`
    `r x3`


```{r, time = TRUE, echo = FALSE}
head(iris)
```

Great!

# Question

Is there any way to see benchmark result right after evaluating chunks in RStudio's editor?
I see the results after knitting.

# Tip

If you are using R Notebook, just preview it and you can see benchmark results.
